Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-27 MScDiplom_Lemenkova.tex
moved 'Data pre-processing' to Methods; changed latin words to Italic font

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to marine environment, they are however, more closely related to the lily family
(Liliaceae) than to the true grass, despite their name “seagrass”, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}.\\
Seagrasses create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass Posidonia oceanica (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

\begin{figure}[h]
	\centering
	\subfloat {\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-1-1a.jpg}}
	\hspace{2mm}
	\subfloat {\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-1-1b.jpg}}
	\caption{Seagrass Posidonia Oceanica}
	\label{fig:1}
\end{figure}

@@

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species (Fig. \ref{fig:2}), belonging to two orders
(Hydrocharitales and Najadales), four families (Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae), and 12 genera (Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera and Phyllospandix) \cite{Kuo89}\label{Kuo89}.

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of P. oceanica are subjected to the human activities, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. One
of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}.

@@

The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (P.
oceanica) in the Mediterranean Sea. Nowadays P.oceanica is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
This naturally increase environmental pressing and leads to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where P.oceanica is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 percent of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.

\subsubsection{General characteristics of Posidonia oceanica}
The endemic Mediterranean seagrass Posidonia oceanica (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

@@

The dominant and most productive coastal ecosystem of the Mediterranean, P.oceanica is
spatially restricted to the Mediterranean area (Fig.\ref{fig:5}), with its extension limited by the western part
of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphological P.oceanica consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots
and short rhizomes (0.5-2.0 mm). The leaves are, perhaps, the most particular characteristics of

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-1-6.jpg}
	\caption{\textit{P.oceanica}.\\
	Source \cite{Luque04}\label{Luque04}}
	\label{fig:6}
\end{wrapfigure}

P.oceanica, making it highly recognizable and distinguishable from other seagrasses (Fig.\ref{fig:6}): having
usual length of 20-40 cm, in some cases they can reach up to 1 m \cite{Borum04}\label{Borum04}  (Fig.\ref{fig:8}).
Growing P.oceanica make a meadows which, in turn, consist of smaller patches, called “matte”
(Fig.\ref{fig:7}), a monumental construction made by the growth of rhizomes and leaves with entangled roots
and entrapped sediment \cite{Francour06}\label{Francour06}.
Representing one of the most productive Mediterranean ecosystems, \textit{P.oceanica} usually serves as a
perfect biological indicator for the assessment of the quality of waters and environmental health
\cite{Boudouresque89}\label{Boudouresque89}. Some authors \cite{Guidetti08,Montefalcone09}\label{Guidetti08} \label{Montefalcone09} used status and population dynamics of \textit{P.oceanica} as indicators for the evaluation

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-1-7.jpg}
	\caption{Scheme of matte structure of \textit{	P.Oceanica}.\\ Source \cite{Pergent90}\label{Pergent90}}
	\label{fig:7}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-1-8.jpg}
	\caption{Structure and components of P.oceanica \\ Source \cite{DiCarlo04}\label{DiCarlo04}}
	\label{fig:8}
\end{wrapfigure}

@@

\subsection{Research objective}
The current MSc research aims to explore the environmental conditions and dynamics of the spatial
distribution of \textit{P.oceanica} seagrass meadows along the Cretan coast, based on the remote sensing and
GIS techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources:
i) spectra of P.oceanica, carbonate sand, silt and other seafloor types
ii) satellite imagery: Landsat, ASTER, MODIS, MERIS
iii) aerial photos: Google Earth
iv) in-situ fieldwork data of underwater videographic measurements
v) vector GIS layers.
The research aims to explore the dynamics of the environmental changes of the P.oceanica meadows
over time (2000-2010), as well as the extent of the spatial distribution over the whole island. The
main research objective is monitoring the seagrass \textit{P.oceanica} along the coasts of Crete, in order to
analyse the environmental changes of the P.oceanica meadows over the last 10 years (2000-2010)
and the spatial extent of its current distribution.
This research is supported by the in-situ measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: underwater videometric measurements made by Olympus camera, aerial
Google Earth and satellite images from different sources, spatial GIS and statistical analysis.

\subsubsection{General objective}
The main objective of this study is to investigate the distribution of seagrass along the northern coast
of Crete using remote sensing techniques.
General objectives:
1) Mapping the spatial distribution of the seagrass \textit{P.oceanica} along the northern coasts of
Crete Island
2) Monitoring the dynamics of the environmental changes of the P.oceanica meadows over
time (2000-2010) in Ligaria beach, Agia Pelagia district.

\subsubsection{Specific objectives}
1. To apply remote sensing data (Google Earth aerial images, Landsat, SPOT satellite images)
for the monitoring of the seagrass meadows distribution
2. To study narrow-band spectral reflectance properties of \textit{P.oceanica} and other seafloor
cover types (sand and silt) using WASI water colour simulation software
3. To use methods of the in situ diving observations and underwater videometric measurements
by Olympus camera in order to receive large-scale imagery of the P.oceanica mattes
4. To perform images classification (supervised and unsupervised) for the thematic mapping of
the P.oceanica seagrass distribution along the coasts of northern Crete.

@@

\subsection{\textit{In-situ} observations of the seagrass meadows}
The traditional methods of in-situ seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03} \label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, markers, GPS and other equipment. The seagrass sampling is
taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the in-situ seagrass observations of the structuring epiphyte
community composition in the P.oceanica ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming in-situ observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}. Other methods of seagrass in-situ monitoring are based on
using the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then
collect the return echoes for the analysis, yet this requires specialized equipment and is mostly used in
the depth waters, combined with bathymetric measurements. Another limitation of the acoustic
techniques is that they are not effective at identifying the composition or even the presence of
submerged aquatic vegetation such as seagrasses and seaweeds \cite{Werdell03}\label{Werdell03}, because
initially acoustic methods were intended for the bathymetric surveying, i.e. they are not adjusted for
the benthic habitat discrimination. The results of the in-situ measurements and observation are usually
managed and treated using integrated GIS approach, as, for example, reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.

@@

Another well-known multispectral sensor, SPOT provides multispectral imagery with a spatial
resolution of 10 m, covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \href{http://www.spotimage.fr/}{SPOT} imagery was used for mapping beds of Posidonia oceanica in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility. Thus,
the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, Landsat TM, and CASI, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and CASI, while Landsat TM has not high
enough resolution. It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that CASI enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.
Another comparative analysis of the application of \href{http://www.itres.com/products/imagers/casi550}{CASI}, Landsat and \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird} imagery \cite{Phinn08}\label{Phinn08}
 demonstrates the high suitability of CASI images for the fine-scale mapping of seagrass
landscapes. Thus, CASI and Quickbird-2 images enable to identify even separate seagrass species
with small width and heterogeneous nature of the seagrass patches, which could not be detected using
Landsat TM images with their 30*30m resolution.

@@

In the section \nameref{sec:2} we briefly discussed main existing cartographic methods and tools for seagrass mapping. In the current  research we try to combine various techniques from mentioned above for the assessment of the seagrass distribution in the Mediterranean environment, for the case study of Crete. We mentioned in \nameref{sec:0} the software that we used for logically separated research sections, which required various approach. Thus, for example,  spectral analysis and assessment of spectral signatures is technically based on WASI, statistical analysis of series of experimental raw data has been made by means of statistical software, such as \href{http://gretl.sourceforge.net/}{Gretl} and \href{http://www.spss.com/}{SPSS}; mapping has been made using  \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS}; images processing and classification were made using \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine} software by means of Google Earth and  \href{http://landsat.gsfc.nasa.gov/}{Landsat} imagery.

@@

The information about the location of the seagrass (mostly represented by the P.oceanica species) is useful for the

@@

Along the Ligaria beach there are several types of the seagrass landscapes, namely, seagrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{fig:A.3} (c)).
The results of the underwater videometric measurements of the Olympus cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass P.oceanica meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig. \ref{fig:A5}). Seagrass species on the Ligaria and Agia
Pelagia beach consist of P.oceanica. The types of sediment on the Ligaria vary from coarse sand,
P.oceanica patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).

\subsection{Data of spectral measurements}

>>>>

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to marine environment, they are however, more closely related to the lily family
(Liliaceae) than to the true grass, despite their name “seagrass”, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}.\\
Seagrasses create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

\begin{figure}[h]
	\centering
	\subfloat {\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-1-1a.jpg}}
	\hspace{2mm}
	\subfloat {\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-1-1b.jpg}}
	\caption{Seagrass \textit{Posidonia Oceanica}}
	\label{fig:1}
\end{figure}

@@

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species (Fig. \ref{fig:2}), belonging to two orders
(\textit{Hydrocharitales} and \textit{Najadales}), four families (\textit{Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae}), and 12 genera (\textit{Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera} and \textit{Phyllospandix}) \cite{Kuo89}\label{Kuo89}.

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. One
of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}.

@@

The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays P.oceanica is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
This naturally increase environmental pressing and leads to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 percent of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.

\subsubsection{General characteristics of \textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

@@

The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area (Fig.\ref{fig:5}), with its extension limited by the western part
of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphological \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots
and short rhizomes (0.5-2.0 mm). The leaves are, perhaps, the most particular characteristics of

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-1-6.jpg}
	\caption{\textit{	extit{P.oceanica}}.\\
	Source \cite{Luque04}\label{Luque04}}
	\label{fig:6}
\end{wrapfigure}

\textit{P.oceanica}, making it highly recognizable and distinguishable from other seagrasses (Fig.\ref{fig:6}): having
usual length of 20-40 cm, in some cases they can reach up to 1 m \cite{Borum04}\label{Borum04}  (Fig.\ref{fig:8}).
Growing \textit{P.oceanica} make a meadows which, in turn, consist of smaller patches, called “matte”
(Fig.\ref{fig:7}), a monumental construction made by the growth of rhizomes and leaves with entangled roots
and entrapped sediment \cite{Francour06}\label{Francour06}.
Representing one of the most productive Mediterranean ecosystems, \textit{P.oceanica} usually serves as a
perfect biological indicator for the assessment of the quality of waters and environmental health
\cite{Boudouresque89}\label{Boudouresque89}. Some authors \cite{Guidetti08,Montefalcone09}\label{Guidetti08} \label{Montefalcone09} used status and population dynamics of \textit{P.oceanica} as indicators for the evaluation

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-1-7.jpg}
	\caption{Scheme of matte structure of \textit{	P.oceanica}.\\ Source \cite{Pergent90}\label{Pergent90}}
	\label{fig:7}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-1-8.jpg}
	\caption{Structure and components of \textit{P.oceanica} \\ Source \cite{DiCarlo04}\label{DiCarlo04}}
	\label{fig:8}
\end{wrapfigure}

@@

\subsection{Research objective}
The current MSc research aims to explore the environmental conditions and dynamics of the spatial
distribution of \textit{P.oceanica} seagrass meadows along the Cretan coast, based on the remote sensing and
GIS techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources:
i) spectra of 	extit{P.oceanica}, carbonate sand, silt and other seafloor types
ii) satellite imagery: Landsat, ASTER, MODIS, MERIS
iii) aerial photos: Google Earth
iv) in-situ fieldwork data of underwater videographic measurements
v) vector GIS layers.
The research aims to explore the dynamics of the environmental changes of the \textit{P.oceanica} meadows
over time (2000-2010), as well as the extent of the spatial distribution over the whole island. The
main research objective is monitoring the seagrass \textit{P.oceanica} along the coasts of Crete, in order to
analyse the environmental changes of the 	extit{P.oceanica} meadows over the last 10 years (2000-2010)
and the spatial extent of its current distribution.
This research is supported by the in-situ measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: underwater videometric measurements made by Olympus camera, aerial
Google Earth and satellite images from different sources, spatial GIS and statistical analysis.

\subsubsection{General objective}
The main objective of this study is to investigate the distribution of seagrass along the northern coast
of Crete using remote sensing techniques.
General objectives:
1) Mapping the spatial distribution of the seagrass \textit{P.oceanica} along the northern coasts of
Crete Island
2) Monitoring the dynamics of the environmental changes of the \textit{P.oceanica} meadows over
time (2000-2010) in Ligaria beach, Agia Pelagia district.

\subsubsection{Specific objectives}
1. To apply remote sensing data (Google Earth aerial images, Landsat, SPOT satellite images)
for the monitoring of the seagrass meadows distribution
2. To study narrow-band spectral reflectance properties of \textit{P.oceanica} and other seafloor
cover types (sand and silt) using WASI water colour simulation software
3. To use methods of the in situ diving observations and underwater videometric measurements
by Olympus camera in order to receive large-scale imagery of the 	extit{P.oceanica} mattes
4. To perform images classification (supervised and unsupervised) for the thematic mapping of
the 	extit{P.oceanica} seagrass distribution along the coasts of northern Crete.

@@

\subsection{\textit{In-situ} observations of the seagrass meadows}
The traditional methods of in-situ seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03} \label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, markers, GPS and other equipment. The seagrass sampling is
taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the in-situ seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming in-situ observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}. Other methods of seagrass in-situ monitoring are based on
using the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then
collect the return echoes for the analysis, yet this requires specialized equipment and is mostly used in
the depth waters, combined with bathymetric measurements. Another limitation of the acoustic
techniques is that they are not effective at identifying the composition or even the presence of
submerged aquatic vegetation such as seagrasses and seaweeds \cite{Werdell03}\label{Werdell03}, because
initially acoustic methods were intended for the bathymetric surveying, i.e. they are not adjusted for
the benthic habitat discrimination. The results of the in-situ measurements and observation are usually
managed and treated using integrated GIS approach, as, for example, reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.

@@

Another well-known multispectral sensor, SPOT provides multispectral imagery with a spatial
resolution of 10 m, covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \href{http://www.spotimage.fr/}{SPOT} imagery was used for mapping beds of \textit{P.oceanica} in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility. Thus,
the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, Landsat TM, and CASI, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and CASI, while Landsat TM has not high
enough resolution. It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that CASI enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.
Another comparative analysis of the application of \href{http://www.itres.com/products/imagers/casi550}{CASI}, Landsat and \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird} imagery \cite{Phinn08}\label{Phinn08}
 demonstrates the high suitability of CASI images for the fine-scale mapping of seagrass
landscapes. Thus, CASI and Quickbird-2 images enable to identify even separate seagrass species
with small width and heterogeneous nature of the seagrass patches, which could not be detected using
Landsat TM images with their 30*30m resolution.

@@

In the section \nameref{sec:2} we briefly discussed main existing cartographic methods and tools for seagrass mapping. In the current  research we try to combine various techniques from mentioned above for the assessment of the seagrass distribution in the Mediterranean environment, for the case study of Crete. We mentioned in the \nameref{sec:0} the software that we used for logically separated research sections, which required various approach. Thus, for example,  spectral analysis and assessment of spectral signatures is technically based on WASI, statistical analysis of series of experimental raw data has been made by means of statistical software, such as \href{http://gretl.sourceforge.net/}{Gretl} and \href{http://www.spss.com/}{SPSS}; mapping has been made using  \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS}; images processing and classification were made using \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine} software by means of Google Earth and  \href{http://landsat.gsfc.nasa.gov/}{Landsat} imagery.

@@

The information about the location of the seagrass (mostly represented by the 	extit{P.oceanica} species) is useful for the

@@

Along the Ligaria beach there are several types of the seagrass landscapes, namely, seagrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{fig:A.3} (c)).
The results of the underwater videometric measurements of the Olympus cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass \textit{P.oceanica} meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig. \ref{fig:A5}). Seagrass species on the Ligaria and Agia
Pelagia beach consist of 	extit{P.oceanica}. The types of sediment on the Ligaria vary from coarse sand,
\textit{P.oceanica} patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).
